---
title: "ðŸ“¦ <br>Building tidy tools"
subtitle: "Day 1 Session 3: Unit tests"
author: "Emma Rand and Ian Lyttle"
date: "<br>ðŸ”— [bit.ly/...](https://bit.ly/..)"
format:
  revealjs: 
    theme: [simple, emma.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: <https://bit.ly/..>
    margin: 0.07
    code-link: true
    code-line-numbers: false
execute:
  echo: true
  eval: false
bibliography: references.bib
---

# â˜‘ï¸ Unit Testing

## Learning Objectives {.smaller}

At the end of this section you will be able to:

-   lo
-   lo
-   lo
-   lo
-   

# Why test?

## Why test?

We already did informal testing:

-   wrote `uss_make_matches()`
-   loaded package with `devtools::load_all()`
-   ran `uss_make_matches()` interactively
-   edited `uss_make_matches()`
-   loaded package with `devtools::load_all()`
-   ran `uss_make_matches()` interactively

## Informal test workflow

```{mermaid}
%%| evaluate:: true
%%| fig-width: 10
%%| fig-height: 5
%%{init: {'theme': 'base', 'themeVariables': { 'primaryColor': '#f3f3f3',  'lineColor':'#EE9AD9'}}}%%

  flowchart LR
      id1("Reload code: \n load_all()") --> 
      id2("Explore in \n console") --> 
      id3("Modify \n code")
      id3 --> id1
      style id1 id2 id3 stroke:#3F3F3F,stroke-width:2px

```

## Why automate testing?

Problem: you forget all the interactive testing you've done

. . .

Solution: have a system to store and re-run the tests!

## Why automate testing?

1.  Fewer bugs: you are explicit about behaviour of functions.

2.  Encourages good code design. If it is hard to write unit tests your function may need refactoring

3.  Opportunity for test-drive development

4.  Robustness

## Automated test workflow

```{mermaid}
%%| evaluate:: true
%%| fig-width: 10
%%| fig-height: 5
%%{init: {'theme': 'base', 'themeVariables': { 'primaryColor': '#f3f3f3',  'lineColor':'#EE9AD9'}}}%%

  flowchart LR
      id1("Reload code: \n load_all()") --> 
      id2("Run automated tests: \n test() or test_file()") --> 
      id3("Modify \n code")
      id3 --> id1
      style id1 id2 id3 stroke:#3F3F3F,stroke-width:2px

```

# Infrastructure and organisation

## Organisation: files {.smaller}

::: columns
::: {.column width="65%"}
``` default
.
â”œâ”€â”€ CITATION.cff
â”œâ”€â”€ DESCRIPTION
â”œâ”€â”€ LICENSE
â”œâ”€â”€ LICENSE.md
â”œâ”€â”€ man
â”‚   â””â”€â”€ matches.Rd
â”œâ”€â”€ NAMESPACE
â”œâ”€â”€ R
â”‚   â””â”€â”€ matches.R
â”œâ”€â”€ README.md
â”œâ”€â”€ tests
â”‚   â”œâ”€â”€ testthat
â”‚   â”‚   â””â”€â”€ test-matches.R
â”‚   â””â”€â”€ testthat.R
â””â”€â”€ ussie.Rproj
```
:::

::: {.column width="35%"}
-   `tests/testthat/`: All test files must be in here
-   `tests/testthat/test-matches.R`: test files must be `test-xxxxx.R`
-   `tests/testthat.R`: runs the tests when `devtools::check()` is called
:::
:::

## Organisation within files

-   any test file `test-xxxxx.R` contains several tests. Might be:
    -   all the tests for a simple function
    -   all the tests for one part of a complex function
    -   all the tests for the same functionality in multiple functions

. . .

-   a test groups several 'expectations'. An expectation:
    -   has the form: `expect_xxxxxxx(actual_result, expectation)`
    -   if `actual_result` = `expectation` no error
    -   if `actual_result` != `expectation` Error

# Workflow

## Workflow

1.  Set up your package to use {testthat} - Once

2.  Make a test

3.  Run the test

# Set up

## Set up

To set up your package to use testthat: `usethis::use_testthat(3)` which:

-   makes `tests/testthat/`: this is where the test files live

-   edits DESCRIPTION: Adds `Suggests:   testthat (>= 3.0.0)` and `Config/testthat/edition: 3`

-   makes `tests/testthat.R`: this runs the test when you do `devtools:check()` DO NOT EDIT

## Set up

ðŸŽ¬ Set up your package to use {`testthat`}:

```{r}
usethis::use_testthat(3)
```

. . .

3 means {`testthat`} edition 3 or {`testthat 3e`}

As well as installing that version of the package, you have to explicitly opt in the edition behaviours.

# Expectations

## Expectations

Before we try to make a test, let's look at some of the `expect_xxxx()` functions we have available to us.

Their form is `expect_xxxx(actual_result, expectation)`

-   the `expectation` is what you expect
-   the `actual_result` is what you are comparing to the expectation
-   some `expect_xxxx()` have additional arguments

## For example

```{r}
#| eval: true
#| error: true

library(testthat)
# when the actual result is 42
result <- 42

# and we expect the result to be 42: no error
expect_identical(result, 42)
```

```{r}
#| eval: true
#| error: true
# when the actual result is "a"
result <- "a"

# and we expect the result to be "a": no error
expect_identical(result, "a")
```

## For example

```{r}
#| eval: true
#| error: true
# when the actual result is 45
result <- 45

# and we expect the result to be 42: Error
expect_identical(result, 42)
```

```{r}
#| eval: true
#| error: true
# when the actual result is "bob"
result <- "bob"

# and we expect the result to be "a": Error
expect_identical(result, "A")
```

## Some common expectations

-   Testing for equality with with wiggle room
-   Testing matches in strings
-   Testing something is TRUE
-   Testing whether objects have names
-   Testing errors

## Testing for equality with wiggle room

```{r}
#| eval: true
#| error: true
# when the actual result is 42
result <- 42

# and we expect the result to be 42: no error
expect_equal(result, 42)
```

```{r}
#| eval: true
#| error: true
# but when the actual result is 42.0000001
result <- 42.0000001

# and we expect the result to be 42: no error
expect_equal(result, 42)
```

```{r}
#| eval: true
#| error: true
# but when the actual result is 42.0000001
result <- 42.1

# and we expect the result to be 42: no error
expect_equal(result, 42, tolerance = 0.1)
```

## Testing matches in strings

## Testing something is TRUE

## Testing whether objects have names

## Testing errors

# Making a test

## Make a test

# Running a test

## Running a test

# Other types of testing

## Snapshot tests

tomorrow!

## 

## 

## 

# 

## 

## 

## 

## 

# 

# â˜‘ï¸ Woo hoo â˜‘ï¸<BR> You .....

## Summary {.smaller}

-   ....
    -   ....
-   .
    -   ...ed

## Summary continued {.smaller}

-   ...
    -   .
-   

## References {.smaller}
