---
title: "📦 <br>Building tidy tools"
subtitle: "Day 2 Session 4: Miscellany"
author: "Emma Rand and Ian Lyttle"
date: "<br>🔗 [rstd.io/btt](https://rstd.io/btt)"
format:
  revealjs: 
    theme: [simple, emma.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: <https://rstd.io/btt>
    margin: 0.07
    code-link: true
    code-line-numbers: false
execute:
  echo: true
  eval: false
bibliography: references.bib
editor: source
---

## Learning objectives 
### Time permitting

At the end of this section you will be able to:

- accept a function an argument to a function using a formula (purrr-style) [@purrr]
- create an S3 class using {vctrs} [@vctrs]
  - recognize that implementing a `format()` method gets you a long way 
  
We will also recap all of today's material and discuss.  
  
## Functions as arguments

Using functions as arguments is a foundation of functional programming.

```r
library("purrr")

ten <- seq(1, 10)
square <- function(x) {
  x^2
}

map_dbl(ten, square)
```

## Anonymous (lambda) functions

Sometimes you want to provide an "on-the-spot" function:

- write it out the long way:

  ```r
  map_dbl(ten, function(x) x^2)
  ```

. . .

- {purrr} interprets one-sided formula using  `.x`:
 
  ```r
  map_dbl(ten, ~.x^2)
  ```

. . .

- R (> 4.1) provides shorthand:

  ```r
  map_dbl(ten, \(x) x^2)
  ```

## Make {purrr} formula into function

```r
my_function <- function(data, f) {
  f <- rlang::as_function(f)
  
  ...
}
```

**Note**:  we cannot *pass the dots* here.

. . .

These are equivalent:

```r
my_function(ten, function(x) x + 1)
```

```r
my_function(ten, \(x) x + 1)
```

```r
my_function(ten, ~.x + 1)
```

## Our turn `"2.4.1"`

. . .

<hr>
**Only** *if needed*, `btt22::btt_reset_hard("2.4.1")`
<hr>

. . . 

*No* new files.

## `uss_make_seasons_final()` 

```r
italy_teams_matches <- 
  uss_get_matches("italy") |> 
  uss_make_teams_matches()
```

What would a season look like if wins were **five** points?

```r
italy_five <-
  italy_teams_matches |>
  uss_make_seasons_final(fn_points_per_win = \(...) 5)
```

## `seasons.R`

 - in `seasons_intermediate()`:

  ```r
  # accept purrr-style anonymous functions
  fn_points_per_win <- 
    rlang::as_function(fn_points_per_win)
  ```

. . .

- `devtools::load()`

  ```r
  italy_five_purrr <-
    italy_teams_matches |>
    uss_make_seasons_final(fn_points_per_win = ~5)
  ```
  
  ```r
  identical(italy_five, italy_five_purrr)
  ```

. . .

- document, `devtools::check()`

## S3 class system

There have been a lot of class systems implemented in R:

- S3, S4, R6, ...

. . . 
 
- and now R7 (Hadley's talk Thursday afternoon)

. . .

For now, S3 remains the simplest thing that works:

 - declare that an object has a class
 - define methods for generic functions, specific to that class, e.g. `format()`
 
## {vctrs} package

- designed to help you construct S3 classes:

  - especially in the context of a column of a data frame

. . . 

- `cli::cli_abort()` creates an S3 error condition:

  - not quite the same thing
 
  - but the same idea: let's make it a little easier to manage

## Our turn `"2.4.2"`

. . .

<hr>
**Only** *if needed*, `btt22::btt_reset_hard("2.4.2")`
<hr>

. . . 
 
Get new files, `btt22::btt_get("2.4.2")`:

- `result.R`
- `test-result.R`

. . . 

- `usethis::use_package("glue")`

- `usethis::use_package("vctrs")`

- add `#' @import vctrs` to `ussie-package.R`.

## What problem are we solving?

A *result* displays two sets of information, according to team:

  - goals-for, goals-against: essential information
  - outcome (win, loss, draw): based on goals for, against
  
. . .

Our *result* structure should:

  - store essential information, e.g.:
    - `goals_for = 3`
    - `goals_against = 2`
  - display outcome with goals, e.g.: `W 3-2`.


## Low-level constructor

```r
new_result <- function(goals_for = integer(), goals_against = integer()) {

  # validate
  vec_assert(goals_for, integer())
  vec_assert(goals_against, integer(), size = length(goals_for))
  
  # construct rcrd class, using vctrs::new_rcrd()
  new_rcrd(
    list(goals_for = goals_for, goals_against = goals_against),
    class = "ussie_result"
  )
}
```

`rcrd` class is like a little data-frame.


## High-level constructor

```r
#' @export
uss_result <- function(goals_for = integer(), goals_against = integer()) {

  # coerce to integer, lets us accept integer-ish arguments
  goals_for <- vec_cast(goals_for, to = integer())
  goals_against <- vec_cast(goals_against, to = integer())
  
  # call low-level constructor
  new_result(goals_for, goals_against)
}
```

This is what users "see" and use.

## Formatter

```r
#' @export
format.ussie_result <- function(x, ...) {
  
  # field() is a vcrts function to extract a field from a rcrd
  goals_for <- field(x, "goals_for")
  goals_against <- field(x, "goals_against")
  
  outcome <- dplyr::case_when(...)
  
  # compose output
  out <- glue::glue("{outcome} {goals_for}-{goals_against}")
  as.character(out)
}
```

## Let's see this in action

- `devtools::document()`
- `devtools::load()`

. . .

```r
uss_result(3, 2)
```

. . .

```r
uss_get_matches("italy") |>
  uss_make_teams_matches() |>
  dplyr::mutate(
    result = uss_result(goals_for, goals_against)
  )
```

## Abbreviators

These control how the class name is displayed when printing:

```r
#' @export
vec_ptype_abbr.ussie_result <- 
  function(x, ...) "rslt" 

#' @export
vec_ptype_full.ussie_result <- 
  function(x, ...) "result"
```

## Recap: Design

- **Naming**: be consistent, concise, yet evocative.

- **Argument order**: data, descriptors, dots, details.


- **Return value type**: be consistent, predictable.


- Easy to remember data (first) argument and return value:
  - easy to use pipe, `|>`.

- Be mindful of side effects.

## Recap: Side-effects

The most-common side-effect is an error. Good design:

- simple (as possible) predicate
- clear mesage
- add a `class` using naming convention, and additional information


Use snapshot tests to capture side-effects.

- be very careful when accepting changes to snapshots.

Use `withr::local_*()` functions to "leave no footprints".

## Recap: Tidy eval

There are a lot of tidy-eval tools:

- know if you are using a *data-masking* or *tidy-select* function.

. . .

- to splice a list (or vector) into *dynamic-dots*, use `!!!`

. . .

- For *data-masking* functions:

  - be specific, use `.data`, `.env` pronouns.
  
  - to use *tidy-select* syntax, use `dplyr::across()`.
  
  - to move a bunch of arguments, pass the dots, `...`
  
  - to interpolate a single argument, use `{{}}`.
  
## Recap: Misc.

Anonymous (use-once) functions can be handy:

  - use `rlang::as_function()` to accept purrr-style formula-functions, e.g. `~.x + 1` 

Use {vctrs} package to help build and manage S3 classes.

## Recap

In Ian's opinion, the tidyverse succeeds because it helps *you* build a mental model for composing functions to solve *your* problem. It does this by providing:

- consistent naming of functions, arguments
- focus on first argument, return value: supports `|>`
- consistent implementation of non-standard evaluation
- more-and-more informative error-messages

## Un-recap

We did not cover how to manage change.

Change is a necessary (and frustrating) part of success:

- {plyr} -> {dplyr}
- {reshape} -> {reshape2} -> {tidyr}

. . .

Different levels of change:

- deprecating functions, arguments
- breaking changes
- used to be: *add-a-2*, becoming: *editions*

## If you remember one thing...

Distinguish *pure functions* from functions that use *side effects*.

Pure functions:

 - use only the values of the arguments
 - change only the return value
 
   ```r
   add <- function(x, y) {
     x + y
   }
   ```

Side effects use or change something else in the universe:

 - e.g.: read a file, print to screen

## Additional material: FP {.smaller}


```{r}
#| eval: true
#| echo: false
vembedr::embed_youtube("e-5obm1G_FY", height = 400) 
```

Anjana Vakil, on functional programming in JavaScript:

- (Ian: JavaScript is a great *next* language from R)
- `filter()` in JavaScript works like `purrr::keep()`

## Additional material: S3 tools

```{r}
#| eval: true
#| echo: false
vembedr::embed_youtube("8_QoT3ygdI4", height = 400) 
```

Hadley Wickham's introduction of {vctrs}, rstudio::conf(2019).

## Additional material: S3 usage

```{r}
#| eval: true
#| echo: false
vembedr::embed_youtube("UHwp8T_fmEI", height = 400) 
```

At rstudio::conf(2020), Jesse Sadler shows an implementation.
 
## References