---
title: "📦 <br>Building tidy tools"
subtitle: "Day 1 Session 2: 🏷️Documentation - Minimal"
author: "Emma Rand and Ian Lyttle"
date: "<br>🔗 [bit.ly/...](https://bit.ly/..)"
format:
  revealjs: 
    theme: [simple, emma.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: <https://bit.ly/..>
    margin: 0.07
    code-link: true
    code-line-numbers: false
execute:
  echo: true
  eval: false
bibliography: references.bib
---

# 🏷️ Document your package

## Learning Objectives {.smaller}

At the end of this section you will be able to:

-   lo
-   lo
-   lo
-   lo
-   

# Overview

## Levels of package documentation

-   Metadata: The `DESCRIPTION` file -- "what's in this package?"

-   Object documentation: for functions and data

<!-- for each of the exported functions and datasets in the package, along with examples of usage -->

-   Vignettes: Long form documentation

<!-- generally discussing how to use a number of functions from the - package together and/or how a package fits into a larger ecosystem of packages -->

-   **`pkgdown`** sites: Websites for your package!

# `DESCRIPTION`

## Sample `DESCRIPTION`

🎬 Take a look at the `DESCRIPTION` for the following packages. What's common? What's different?

-   **`ggplot2`** @ggplot2
    -   [CRAN Package page](https://cran.r-project.org/web/packages/ggplot2/index.html)
    -   [DESCRIPTION on GitHub](https://github.com/tidyverse/ggplot2/blob/main/DESCRIPTION)
-   **`palmerpenguins`** @palmerpenguins
    -   [CRAN Package page](https://cran.r-project.org/web/packages/palmerpenguins/index.html)
    -   [DESCRIPTION on GitHub](https://github.com/allisonhorst/palmerpenguins/blob/master/DESCRIPTION)

## Metadata in `DESCRIPTION`

-   Title: One line, title case, with no period. Fewer than 65 characters.

-   Version

    -   for release: MAJOR.MINOR.PATCH version.
    -   for development version building on version MAJOR.MINOR.PATCH, use: MAJOR.MINOR.PATCH.9000

-   Authors\@R:

    -   "aut" means author, "cre" means creator, "ctb" means contributor.

## Update `DESCRIPTION`

🎬 Add a title

\

🎬 You should find that your information as an author is already included because you edited your `.RProfile` at the start of the session

## Update `DESCRIPTION` {.smaller}

    Package: ussie
    Title: Work with European Football League Data
    Version: 0.0.0.9000
    Authors@R: 
        person(given = "Emma",
               family = "Rand", 
               email = "emma.rand@york.ac.uk",
               role = c("aut", "cre"),
               comment = c(ORCID = "0000-0002-1358-8275"))
    Description: What the package does (one paragraph).
    License: MIT + file LICENSE
    Encoding: UTF-8
    Roxygen: list(markdown = TRUE)
    RoxygenNote: 7.2.0
    URL: https://github.com/3mmaRand/ussies
    BugReports: https://github.com/3mmaRand/ussies/issues



## Metadata in `DESCRIPTION`

-   Description: One paragraph describing what the package does. Keep the width of the paragraph to 80 characters; indent subsequent lines with 4 spaces

-   License

-   Encoding: How to encode text, use UTF-8 encoding.

-   LazyData: Use true to lazy-load data sets in the package.



# Object documentation

## Object documentation

-   Object documentation is what you see when you use `?` or `help()`

-   These helps files are written in a special "R documentation" format: `.Rd`

-   `.Rd` resembles LaTeX,

## `man/uss_make_matches.Rd` {.smaller}

    \name{uss_make_matches}
    \alias{uss_make_matches}
    \title{Make a standard league-play tibble}
    \usage{
    uss_make_matches(data_engsoc, country)
    }
    \arguments{
    \item{data_engsoc}{\code{data.frame} obtained from {engsoccerdata}. One of:
    engsoccerdata::england, engsoccerdata::germany, engsoccerdata::holland,
    engsoccerdata::italy, engsoccerdata::spain}

    \item{country}{\code{character} scalar, specifies the league.}
    }
    \value{
    a tibble with columns \code{country}, \code{date}, \code{season}, \code{tier}, \code{home},
    \code{visitor}, \code{goals_home}, \code{goals_visitor}.
    }
    \description{
    Given a league-play data frame from {engsoccer}, returns a tibble with
    standardised colomn-names and types, e.g. \code{date} is a \code{Date}.
    }
    \examples{
    uss_make_matches(engsoccerdata::italy, "Italy")
    uss_make_matches(engsoccerdata::spain, "Spain")

    }

## Object documentation

🥳 We don't have to write it!!

. . .

-   We write "roxygen comments" in the `.R` files

## Roxygen comments

    #' Make a standard league-play tibble
    #'
    #' Given a league-play data frame from {engsoccer}, returns a tibble with
    #' standardised colomn-names and types, e.g. `date` is a `Date`.
    #'
    #' @param data_engsoc `data.frame` obtained from {engsoccerdata}. One of:
    #'   engsoccerdata::england, engsoccerdata::germany, engsoccerdata::holland,
    #'   engsoccerdata::italy, engsoccerdata::spain
    #'
    #' @param country `character` scalar, specifies the league.
    #'
    #' @return a tibble with columns `country`, `date`, `season`, `tier`, `home`,
    #'   `visitor`, `goals_home`, `goals_visitor`.
    #'
    #' @export
    #'
    #' @examples
    #' uss_make_matches(engsoccerdata::italy, "Italy")
    #' uss_make_matches(engsoccerdata::spain, "Spain")
    #'

-   the `#'` indicates it is a roxygen comment

-   **`roxygen2`** does the work but we can use **`devtools`** functions to call them

## Object documentation workflow

-   Add roxygen comments to your `.R` files.

-   Run `devtools::document()` to convert roxygen comments to `.Rd` files.

-   Load the package with `devtools::load_all()`

-   Preview documentation with `?`

-   Repeat until the documentation looks the way you want.

## Document your function

🎬 Open `uss_make_matches.R`

🎬 Put your cursor anywhere in the function and do Code \| Insert Roxygen Documentation (Ctrl-Alt-Shift-R)

## Document your function {.smaller}

    #' Title
    #'
    #' @param data_engsoc 
    #' @param country 
    #'
    #' @return
    #' @export
    #'
    #' @examples
    uss_make_matches <- function(data_engsoc, country) {
      result <-
        data_engsoc |>
        tibble::as_tibble() |>
        dplyr::transmute(
          country = as.character(country),
          tier = as.integer(tier),
          season = as.integer(Season),
          date = as.Date(Date),
          home = as.character(home),
          visitor = as.character(visitor),
          goals_home = as.integer(hgoal),
          goals_visitor = as.integer(vgoal)
        )
      result
    }

## Document your function

-   `#'` indicates it is a roxygen comment

-   `@` indicates a roxygen tag

    -   `@param arg` --- describe the inputs
    -   `@examples`--- show how the function works
    -   `@seealso` --- point out related functions
    -   `@returns` --- describe the outputs
    -   `@param arg` --- is this a user visible function

## Document your function

🎬 Give your function a title and a brief description

🎬 Define the two parameters

🎬 Describe what the function returns

## Document your function {.smaller}

    #' Make a standard league-play tibble
    #'
    #' Given a league-play data frame from {engsoccer}, returns a tibble with
    #' standardised colomn-names and types, e.g. `date` is a `Date`.
    #'
    #' @param data_engsoc `data.frame` obtained from {engsoccerdata}. One of:
    #'   engsoccerdata::england, engsoccerdata::germany, engsoccerdata::holland,
    #'   engsoccerdata::italy, engsoccerdata::spain
    #'
    #' @param country `character` scalar, specifies the league.
    #'
    #' @return a tibble with columns `country`, `date`, `season`, `tier`, `home`,
    #'   `visitor`, `goals_home`, `goals_visitor`.

## Take a look at your documentation

🎬 Save `uss_make_matches.R` and run `devtools::document()`

. . .

    Warning message:
    [uss_make_matches.R:17] @examples requires a value

We will that fix later!

. . .

🎬 Load the package : `devtools::load_all()`

🎬 Preview the documentation with `?uss_make_matches`

## Add examples

🎬 Under `@examples`, add one example for using your function

. . .

    #' @examples
    #' uss_make_matches(engsoccerdata::italy, "Italy")
    #' uss_make_matches(engsoccerdata::spain, "Spain")
    #' 

## Take a look at your documentation

🎬 Save `uss_make_matches.R`, run `devtools::document()` followed by `devtools::load_all()`

. . .

🎬 Preview the documentation with `?uss_make_matches` and edit your documentation if anything needs to be changed

## `check()`

## add another function?

## document it?

## check


# Package dependecies

## Remember this?



```
> checking dependencies in R code ... WARNING
  '::' or ':::' imports not declared from:
    'dplyr' 'tibble'

> checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata''
```


. . . 

We have used packages that we have not declared officially.


## Package dependencies

- **Imports:** must be installed for your package to work. If they're not, they will get installed.

- **Suggests:** used by your package, but not required. Might provide data for examples, to run tests, build vignettes.

- **Depends:** Avoid where possible. When your package requires a specific version of R, e.g. `Depends: R (>= 3.4.0)`. Think critically: downstream effects on packages that depend on your package.

## Package dependencies

`use_package(package, type = "Imports")`

-   Type -- one of "Imports", "Depends", "Suggests", "Enhances", or "LinkingTo" 

-   The default is "Imports"

## `usethis::use_package()`


🎬 Use `usethis::use_package()` to add the **`dplyr`** package to `Imports`

```{r}
usethis::use_package("dplyr")
```

. . .

```
v Setting active project to 'C:/Users/er13/OneDrive - University of York/Desktop/Desktop/ussie'
v Adding 'dplyr' to Imports field in DESCRIPTION
* Refer to functions with `dplyr::fun()`
```

Look how your `DESCRIPTION` file changed!


## `usethis::use_package()`


Note: we get reminded to "Refer to functions with `dplyr::fun()`"

That is, we do not use `library(dplyr)` to make functions available to our package.


. . .


🎬 Repeat for **`tibble`** 

```{r}
usethis::use_package("tibble")
```

## `check()`

🎬 Run `check()` on your package again

## {.smaller}

```
-- R CMD check results -------------------------------- ussiesun 0.0.0.9000 ----
Duration: 53.5s

> checking for unstated dependencies in examples ... WARNING
  '::' or ':::' import not declared from: 'engsoccerdata'

> checking R code for possible problems ... NOTE
  uss_make_matches: no visible binding for global variable 'tier'
  uss_make_matches: no visible binding for global variable 'Season'
  uss_make_matches: no visible binding for global variable 'Date'
  uss_make_matches: no visible binding for global variable 'home'
  uss_make_matches: no visible binding for global variable 'visitor'
  uss_make_matches: no visible binding for global variable 'hgoal'
  uss_make_matches: no visible binding for global variable 'vgoal'
  Undefined global functions or variables:
    Date Season hgoal home tier vgoal visitor

0 errors v | 1 warning x | 1 note x
```

## Suggests

-   Are used by your package, but not required for it to work


-   Might provide data for examples


`uss_make_matches.R`

```
#' @examples
#' uss_make_matches(engsoccerdata::italy, "Italy")
#' uss_make_matches(engsoccerdata::spain, "Spain")

```

## Suggests


🎬 Add **`engsoccerdata`** 

```{r}
usethis::use_package("engsoccerdata", type = "Suggests")
```

```
v Setting active project to 'C:/Users/er13/OneDrive - University of York/Desktop/Desktop/ussie'
v Adding 'engsoccerdata' to Suggests field in DESCRIPTION
* Use `requireNamespace("engsoccerdata", quietly = TRUE)` to test if package is installed
* Then directly refer to functions with `engsoccerdata::fun()`

```


## Suggests

-   Packages listed in Suggests are not automatically installed along with your package. 

-   This means that you can’t assume the package is available

```
* Use `requireNamespace("engsoccerdata", quietly = TRUE)` to test if package is installed
```


## CITATION.cff

https://docs.ropensci.org/cffr/

```{r}
cffr::cff_write()

```

. . .

```
CITATION.cff generated
Adding  CITATION.cff to .Rbuildignore

cff_validate results-----
Congratulations! This .cff file is valid
```

# 🏷️Yee haw! 🏷️ <BR> Your package has has documentation!

## Summary {.smaller}

-   .
    -   .
    -   .
-   .
    -   .

## Summary continued {.smaller}

-   

-   We add documentation using **`roxygen2`** and `devtools::document()`

## References {.smaller}
