---
title: "📦 <br>Building tidy tools"
subtitle: "Day 2 Session 2: Managing Side Effects"
author: "Emma Rand and Ian Lyttle"
date: "<br>🔗 [rstd.io/btt](https://rstd.io/btt)"
format:
  revealjs: 
    theme: [simple, emma.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    footer: <https://rstd.io/btt>
    margin: 0.07
    code-link: true
    code-line-numbers: false
execute:
  echo: true
  eval: false
bibliography: references.bib
editor: source
---

## Learning objectives 

At the end of this section you will be able to:

- write effective error messages
- create validator functions
- manage the global state (leave no footprints) using {withr}
- test functions that use or cause side-effects
- use snapshot testing

## Errors

- The most common type of side-effect is the **error condition**. 

. . .

- Sometimes, error messages can be cryptic: 

  ```r
  seq[10]
  ```

  ```
  Error in seq[10] : object of type 'closure' is not subsettable
  ```

. . . 


- You can write error messages that make things clearer for: 
  - developers who call your functions
  - end users

## Creating an error condition 

An effective error condition has:

- **predicate** (logical expression used to identify condition)
- clear **message** for end user
- **class name** for developer
- **more information** for developer

## Using `cli::cli_abort()`

{cli} package [@cli]

```r
# predicate
if (y > 3) {
  cli::cli_abort(
    # message for end user
    c(
      "{.var y} cannot be greater than 3.", 
      x = "{.var y} is {.val {y}}."
    ),
    # class name
    class = "ussie_error_threshold",  
    # more information
    y = y     
  )
}
```

## Predicate

Prefer:

 - simpler predicates and more error-conditions
 
Over:

 - complex predicates and fewer error-conditions
 
. . .

Finding simplest set of predicates is just as challenging as finding the "right" names for functions and arguments.

## Message

Content:

- how did we violate the predicate?

. . . 

Formatting:

- {cli} provides powerful formatting tools:

  - use curly-braces and a tag, e.g. `{.var y}` 
  - use more curly-braces to interpolate, e.g. `{.val {y}}`
  
- see [cli inline-markup](https://cli.r-lib.org/reference/inline-markup.html) for more details.

## Class name

- A developer, calling your function, can use the `class` name to handle the error, if they want.

- Convention: 
  
  - `"{package}_error_{description}"`

## Additional information

This "stuff" is also available to an error handler.

- Provide the data that went into the predicate.

- Provide name of variable, e.g. `y = y`.

- Avoid reserved names: `message`, `class`, `call`, `body`, `footer`, `trace`, `parent`, `use_cli_format`.

## Validation

If there will be an error, surface it quickly.

Validating the arguments to a function is one way to do this.

. . .

For example:

 - is *this* a data frame?
 - does this data frame have *these* columns?

. . . 
 
Questions like these can be generalized into functions:

 - throw an error if you need to.
 - otherwise, return **data** argument invisibly.
 
## Validate string-values 

You may be familiar with `match.arg()`:

 - uses "magic" to compare value (if any) to default
 - argument default is vector of strings
   - if value among defaults, return it - otherwise **error**
   - if no value, return *first* value in default

. . .

`rlang::arg_match()` does the same thing, but:

  - partial match triggers error
  - error messages conform to tidyverse standards

## Snapshot tests

Designed for capturing side-effects:

 - error messages
 - `dplyr::glimpse()` is a useful shortcut
 
. . .  
 
Be careful about accepting changes (don't just accept).

Can be temperamental - not run on CRAN.

We will go through, using examples.
   
## Your turn `"2.2.1"`

Implement validator-functions:

 - is the `country` valid?
 
. . .

<hr>
**Only** *if needed*, `btt22::btt_reset_hard("2.2.1")`
<hr>

. . . 
 
Get new files, `btt22::btt_get("2.2.1")`:

 - `validate.R`
 - `test-validate.R`

`usethis::use_package("cli")`

## Use `rlang::arg_match()` {.smaller}

👉 `devtools::load_all()`, then:

- 👉 `uss_get_matches("italy")` 🎉
- 👉 `uss_get_matches("tatooine")` 🎉 
- 👉 `uss_get_matches("ita")` 🤔

. . .

<hr>


`get-matches.R`: 

 - replace `match.arg()` with `rlang::arg_match()`
 
`test-get-matches.R`: 

 - uncomment test for `2.2.1`, then repeat 👉

## Build error constructor {.smaller}

- `usethis::use_package("cli")`

- review `validate_data_frame()` (`call` argument)

- build constructor for `validate_cols()`:
  - something like `{.field {cols}}` may be useful
  - activate `2.2.1` tests in `test-validate.R`, as you go

- hold off on snapshot test (we'll do together)

. . .

- add validator-functions to `uss_make_matches()`:
  - `matches.R`
  - `cols_engsoc()` may be useful

## Managing global state

Leave no footprints.

Leave the global state how you found it, avoid surprises later:

 - packages loaded
 - also: options, environment variables 

. . .

Loading {devtools} in `.Rprofile` changes the global state.

. . .

When we hit the "Knit" button, or the Quarto "Render" button:

  - it runs in a *new* R session
  - does not execute user's `.Rprofile`
  
## Using "self-removing footprints"

The {withr} package [@withr] gives us tools to:

 - modify global state
 - specify when to reverse the modification

. . . 
 
Useful family of functions:

 - `withr::local_*()`
 - changes the state back when a scope is exited
 - if called within a function, *normally* when the function exits

## When could this be useful?

CRAN is (rightly) particular about "leave no footprints".

You may need:

 - `withr::local_options()`: change an option
 - `withr::local_dir()`: change the working directory
 - `withr::local_tempfile()`: path to a temporary file
 
. . . 

Useful in `testthat` code and in `R` code.

## Interactive example

```{r}
write_read_vanish <- function(x) {
  
  tempfile <- withr::local_tempfile(fileext = ".rds")
  
  saveRDS(x, file = tempfile)
  xnew <- readRDS(tempfile)
  
  print(
    glue::glue("'{tempfile}' contained {xnew}.")
  )
  
  invisible(xnew)
}
```

## Another interactive example

```r
not_runif <- function(n, min = 0, max = 1) {
  withr::local_seed(314)
  
  runif(n, min = min, max = max)
}
```

## Example

- `usethis::use_test("validate")`
- change: 

  ```r
  expect_identical(error_condition$cols_req, "foo")
  ```
  
- to: 

  ```r
  expect_identical(error_condition$cols_re, "foo")
  ```

- rerun tests 🤔

. . . 

- By default, `$` accepts partial matching.

- We want stricter testing, so we need to set some options.

## Our turn `"2.2.2"`

Turn off `$` partial matching when testing.

. . .

<hr>
**Only** *if needed*, `btt22::btt_reset_hard("2.2.2")`
<hr>

. . . 
 
Get new files, `btt22::btt_get("2.2.2")`:

 - `utils-testing.R`
 
`usethis::use_package("withr")` 

## `local_warn_partial_match()`

In essence, this sets (temporarily) `TRUE`:

 - `options("warnPartialMatchDollar")`
 - `options("warnPartialMatchArgs")`
 - `options("warnPartialMatchAttr")`

. . .

With a couple of wrinkles:

 - we need to treat `NULL` as `FALSE`.
 - we set things back when the *calling* scope exits.

## Our turn `"2.2.2"`, continued

- `usethis::use_test("validate")`

- add `local_warn_partial_match()` to top of blocks.

- `getOption("warnPartialMatchDollar")`

- change: 

  ```r
  expect_identical(error_condition$cols_re, "foo")
  ```

- `devtools::test()` 🎉 

- `getOption("warnPartialMatchDollar")`

- change back to `$cols_req`

## Your turn `"2.2.3"` {.smaller}

Implement testing with side effects

. . .

<hr>
**Only** *if needed*, `btt22::btt_reset_hard("2.2.3")`
<hr>

. . . 

*No* new files.

. . .

- add `local_warn_partial_match()` to all test-blocks.

- `test-matches.R`:
  - add `expect_error()` calls for data-frame and columns (use `class` argument).
  - at end, add:
   ```r
   expect_snapshot(dplyr::glimpse(italy))
   ```

## Summary

. . .

The most-common side-effect is an error. Good design:

. . .

- simple (as possible) predicate
- clear mesage
- add a `class` using naming convention, and additional information

. . .

Use snapshot tests to capture side-effects.

- be very careful when accepting changes to snapshots.

. . . 

Use `withr::local_*()` functions to "leave no footprints".

## Additional material {.smaller}

```{r}
#| eval: true
#| echo: false
vembedr::embed_youtube("ZhuHCtR3xq8", height = 400) 
```

Brian Beckman on monads (Ian watched at least ten times, learned something each time):

- tidyverse is a monoidal system

- maybe monads aren't the thing, but the functional-programming learned along the way

## References
