---
title: "Prerequisites"
author: "Emma Rand and Ian Lyttle"
format: html
execute:
  echo: true
  eval: false
bibliography: references.bib
---

You will need

-   [R and RStudio](#install-r-and-rstudio)
-   [R package development toolchain](#install-r-package-development-toolchain): Rtools(windows) or XCode (mac) or r-base-dev
-   To [install the **`devtools`** and **`btt22`** packages](#packages)
-   [Git](#git)
-   [GitHub account](#github-account)
-   [GitHub to be linked to RStudio](#check-link-between-rstudio-and-github)

## Install R and RStudio {#install-r-and-rstudio}

You will need:

-   R 4.x. Version 3.6.x will likely be fine but we recommend that you install the latest versions if possible.
-   RStudio 2022.02.1 or above

### Installing R

Download the pre-compiled binary for your OS from <https://cloud.r-project.org/> and install.

#### For Windows

Click "Download R for Windows", then "base", then "Download R #.#.# for Windows". This will download an `.exe` file; once downloaded, open to start the installation. If you do not have administrator rights, the installer will default to install in your Documents folder - if you prefer, you can change the location to another folder that you have write access to.

#### For Mac

Click "Download R for (Mac) OS X", then "R-#.#.#.pkg" to download the installer. Run the installer to complete installation.

#### For Linux

Click "Download R for Linux". Instructions on installing are given for Debian, Redhat, Suse and Ubuntu distributions. Where there is a choice, install both `r-base` and `r-base-dev`.

### Installing R Studio

Downloads are available from <https://www.rstudio.com/products/rstudio/download/> (scroll to the end of the page to see the downloads). You want RStudio Desktop (Open Source License).

#### For Windows with no admin rights

Download the `.zip` source archive under "Zip/Tarballs". Extract the files to a folder where you have write access, e.g. `C:\Users\username\RStudio`. In this folder, open the `bin` directory and find the RStudio program: it is named `rstudio.exe`, but the file extension will typically be hidden, so look for `rstudio`. Right-click this executable to create a desktop shortcut. Double-click the executable or use the shortcut to open.

#### For all other operating systems

Download the relevant installer for your OS listed under "Installers for Supported Platforms". Run the installer to complete installation.

## Install R package development toolchain {#install-r-package-development-toolchain}

#### For Windows with no admin rights

Download the latest (recommended) Rtools installer, `Rtools.exe`, from <https://cran.r-project.org/bin/windows/Rtools/>. Run the installer keeping the default settings.

#### For Windows with admin rights

Download the latest (recommended) Rtools installer, `Rtools.exe`, from <https://cran.r-project.org/bin/windows/Rtools/>. Run the installer making the following selections: keep the default settings for the installation location and components to install; check the box to add rtools to the system PATH.

#### For Mac

Install XCode. Either:

1.  Download and install XCode from the Mac AppStore: <http://itunes.apple.com/us/app/xcode/id497799835?mt=12>
2.  Within XCode go to Preferences : Downloads and install the Command Line Tools

Or for a smaller download size:

1.  Register as an Apple Developer (free) here: <https://developer.apple.com/programs/register/>
2.  Download the Command Line Tools for XCode appropriate for the version of OS X you are running from here: <https://developer.apple.com/downloads/>

#### For Linux

If you installed `r-base-dev`, when installing R, you should have all you need to build packages from source. Otherwise return to the instructions for installing R for your distribution and follow the instructions there to install the additional tools.

## Install the **`devtools`** and **`btt22`** packages {#packages}

The aim of **`devtools`** [@devtools] is to make package development easier by providing R functions that simplify and expedite common tasks. The [R Packages](https://r-pkgs.org/) book and the Building Tidy Tools workshop are based on a **`devtools`** workflow.

Install **`devtools`**:

```{r}
install.packages("devtools")
```

After installing the [R package development toolchain](#install-r-package-development-toolchain) and **`devtools`**, you can verify your system is set up by running:

```{r}
#| eval: true
devtools::has_devel()
```

The central project for the Building Tidy Tools 2022 workshop is you will build a package called **`ussie`**, and populate it with functions. The goal of **`btt22`** is to help you build the **`ussie`** package. Its two main purposes are:

-   installing **`btt22`** means you will have all the packages you need for the workshop and,
-   **`btt22`** has functions that allow you to template code for workshop exercises and to reset the state of **`ussie`** to a particular point in the workshop if things go wrong.

*We recommend you install **`btt22`** no sooner than a week before the workshop.*

Install **`btt22`** from GitHub:

```{r}
devtools::install_github("rstudio-conf-2022/btt22")
```

## Git {#git}

### Windows

Check if Git is installed by running `which git` in the **Terminal**

If `which git` didn't find Git installed:

-   Download and install from <https://git-scm.com/downloads>
-   Keep all the default settings
    -   If the installer hangs with the progress bar at 100%, close the installer with Task Manager (press Ctrl + Alt + Delete; select Task Manager; find Git for Windows installer and close).
-   Open RStudio. In the menus go to Tools \> Global Options \> Git/SVN.
-   Check RStudio has found git under "Git executable:"
    -   You may need to click Browse and find the git executable.

    -   It may be under C:\\Users\\username\\AppData\\Local\\Programs\\Git\\ - remember if file extensions are hidden it will show as git rather than git.exe. Restart RStudio before trying to use git.

### Mac

Check if Git is installed by running `which git` in the **Terminal**

If asked to install the Xcode command line tools, say yes!

If `which git` didn't find Git installed, and if you weren't prompted to install it, run the following in the **Terminal**: `xcode-select --install`

#### Other options for installing Git on a Mac

-   Recommended if you don't want to use the Terminal:
    -   Download and install from <https://git-scm.com/downloads>

    -   If you're developing version controlled packages, you should get to using the Terminal
-   Recommended if you do lots of scientific computing and using software that needs to be installed and updated regularly:
    -   You might first need to install Homebrew if this is your first time using it, see [brew.sh](https://brew.sh/) for instructions

    -   Use Homebrew: Run the following in the Terminal: `brew install git`

## GitHub account {#github-account}

Register a GitHub account at [github.com](https://github.com/).

-   Incorporate your actual name!
-   Reuse your username from other contexts.
-   Pick a username you will be comfortable revealing to your future boss.
-   Shorter is better than longer.
-   Be as unique as possible in as few characters as possible.
-   Make it timeless.
-   Avoid words laden with special meaning in programming.
-   Use all lower case letters.

Advice and more information from [Happy Git with R](https://happygitwithr.com/) [@bryan]

## Check link between RStudio and GitHub {#check-link-between-rstudio-and-github}

Configure your `user.name` and `user.email` for git in RStudio with:

```{r}

usethis::use_git_config(
  user.name = "Jane Doe",          # actual first and last name
  user.email = "jane@example.com"  # email associated with GitHub account
  )
```

You can check this by asking for a "situation report"

```{r}
usethis::git_sitrep()
```

### Personal Access Token

Communicating with GitHub will require authentication. Instead of typing in your username and password each time, use a PAT (personal access token)

Create a PAT with:

```{r}
usethis::create_github_token()
```

-   Give the token a descriptive name
-   Copy and also store this token somewhere (you won't be able to see it again!)

Now put your PAT into the Git credential store with:

```{r}
gitcreds::gitcreds_set()
```

::: aside
You may need to install **`gitcreds`**.
:::

The [Managing Git(Hub) Credentials](https://usethis.r-lib.org/articles/articles/git-credentials.html) vignette in the usethis package has all the details on PATs!
