---
title: "Prerequisites"
author: "Emma Rand and Ian Lyttle"
execute:
  echo: true
  eval: false
---

You will need

-   [R and RStudio](#install-r-and-rstudio)

-   [R package development toolchain](#install-r-package-development-toolchain): Rtools(windows) or XCode (mac) or r-base-dev

-   The following [packages: devtools and assertthat](#packages)

-   [Git and a GitHub account and verified it is linked to RStudio](#git-and-a-github-account-and-verified-it-is-linked-to-rstudio)

You can verify your system is set up by running:

`devtools::has_devel()`

and

`usethis::git_sitrep()`

## Install R and RStudio {#install-r-and-rstudio}

You will need: - R 4.x. Version 3.6.x will likely be fine but we recommend that you install the latest versions if possible - RStudio 1.2 or above

### Installing R

Download the pre-compiled binary for your OS from https://cloud.r-project.org/ and install. More specifically, for version number #.#.# (e.g. 4.0.0):

#### For Windows

Click "Download R for Windows", then "base", then "Download R #.#.# for Windows". This will download an `.exe` file; once downloaded, open to start the installation. If you do not have administrator rights, the installer will default to install in your Documents folder - if you prefer, you can change the location to another folder that you have write access to.

#### For Mac

Click "Download R for (Mac) OS X", then "R-#.#.#.pkg" to download the installer. Run the installer to complete installation.

#### For Linux

Click "Download R for Linux". Instructions on installing are given for Debian, Redhat, Suse and Ubuntu distributions. Where there is a choice, install both `r-base` and `r-base-dev`.

### Installing R Studio

Downloads are available from https://www.rstudio.com/products/rstudio/download/ (scroll to the end of the page to see the downloads). You want RStudio Desktop (Open Source License).

#### For Windows with no admin rights

Download the `.zip` source archive under "Zip/Tarballs". Extract the files to a folder where you have write access, e.g. `C:\Users\username\RStudio`. In this folder, open the `bin` directory and find the RStudio program: it is named `rstudio.exe`, but the file extension will typically be hidden, so look for `rstudio`. Right-click this executable to create a desktop shortcut. Double-click the executable or use the shortcut to open.

#### For all other operating systems

Download the relevant installer for your OS listed under "Installers for Supported Platforms". Run the installer to complete installation.

## Install R package development toolchain {#install-r-package-development-toolchain}

#### For Windows with no admin rights

Download the latest (recommended) Rtools installer, `Rtools.exe`, from <https://cran.r-project.org/bin/windows/Rtools/>. Run the installer keeping the default settings.

#### For Windows with admin rights

Download the latest (recommended) Rtools installer, `Rtools.exe`, from <https://cran.r-project.org/bin/windows/Rtools/>. Run the installer making the following selections: keep the default settings for the installation location and components to install; check the box to add rtools to the system PATH.

#### For Mac

Install XCode. Either:

1.  Download and install XCode from the Mac AppStore: <http://itunes.apple.com/us/app/xcode/id497799835?mt=12>

2.  Within XCode go to Preferences : Downloads and install the Command Line Tools

Or for a smaller download size:

1.  Register as an Apple Developer (free) here: <https://developer.apple.com/programs/register/>

2.  Download the Command Line Tools for XCode appropriate for the version of OS X you are running from here: <https://developer.apple.com/downloads/>

#### For Linux

If you installed `r-base-dev`, when installing R, you should have all you need to build packages from source. Otherwise return to the instructions for installing R for your distribution and follow the instructions there to install the additional tools.

## Packages {#packages}

Install `devtools`

```{r}
install.packages("devtools")
```

Install `assertthat`

```{r}
install.packages("assertthat")
```

## Git and a GitHub account and verified it is linked to RStudio {#git-and-a-github-account-and-verified-it-is-linked-to-rstudio}

### Git

### GitHub account

### Check link between RStudio and GitHub
